---
output:
  pdf_document: default
  html_document: default
---
# Supplementary Material

Horne _et. al._, _"Wechsler Test of Adult Reading in Parkinson’s: a stable yet problematic measure of premorbid cognitive function."_

Here we report the analysis of the above paper against the following checklist 
for reporting latent class trajectory analyses:

van de Schoot, R., Sijbrandij, M., Winter, S. D., Depaoli, S., & Vermunt, J. K. 
(2017). The GRoLTS-checklist: Guidelines for reporting on latent trajectory 
studies. _Structural Equation Modeling: A Multidisciplinary Journal,_ ***24,*** 
451–467. https://doi.org/10.1080/10705511.2016.1247646


## 1. Is the metric of time used in the statistical model reported?

We used age as the time metric (in years to one decimal place). This was to 
allow for direct comparison against controls (which would not be possible with 
disease duration). Follow-up time within the study was also not considered, as 
participants can enter the study at arbitrary points (in terms of both age and 
disease duration).

To avoid instabilities in the cubic models, age was standardised by subtracting 
the global mean age at assessment and dividing by ten. For comparability, this 
was also done for the linear models. In the visualisations, predictions were 
projected back into untransformed age in years.

## 2. Is information presented about the mean and variance of time within a wave?

The metric was age rather than intended study time points. The overall goal in 
the study was to follow-up participants two-yearly, however, the interval 
between follow-up times was multi-modal and varied both within and across 
participants (see Figure \@ref(fig:fig-A1-wtar-interval-histogram)). This 
variability was due to both changing study demands (such as recruiting 
participants for sub-studies) and participant health and availability.

```{r fig-A1-wtar-interval-histogram, fig.cap='Frequencies of the interval between successive WTAR assessments.', fig.width = 12/2.54, fig.height = 10/2.54}
dat %>%  
  filter(!is.na(wtar_interval)) %>% # no interval in first session
  ggplot(aes(x = wtar_interval)) +
  labs(title = 'Interval in years between successive WTAR assessments',
       x = NULL, y = NULL) +
  scale_x_continuous(breaks = seq(from = 0, to = 10, by = 2),
                     labels = scales::number_format(accuracy = 1, 
                                                    suffix = ' years')) +
  geom_histogram(binwidth = 0.1, colour = 'white', size = 0.2) + 
  theme_bw() +
  theme(plot.title = element_text(face = 'bold', size = 10),
        axis.ticks = element_line(colour = '#E8E8E8'),
        panel.border = element_blank())

```

## 3a. Is the missing data mechanism reported?

Data was not missing at random: participants are more likely to drop out of the 
study due to increasing disease severity, other health issues, and old age. As 
shown in Figure \@ref(fig:figure-2-trajectories), however, many remained in the 
study until the end point of WTAR collection (the onset of dementia). Those 
participants did not show a decline in WTAR performance relative to other 
participants. Due to the findings of this study (that the WTAR is relatively 
stable throughout the disease course), differential drop-out is unlikely to 
affect the results.

## 3b. Is a description provided of what variables are related to attrition/missing data?

They are speculated upon in _3a_ above.

## 3c. Is a description provided of how missing data in the analyses were dealt with?

As noted in _3a_ above, missing data was not thought to be problematic in this 
analysis and was not included explicitly in the modeling process.

## 4. Is information about the distribution of the observed variables included?

In Figure \@ref(fig:figure-1-histogram) in the accompanying manuscript, we show 
the distribution of the WTAR-estimated IQ scores in each group, showing the 
cognitive status at each observation.

## 5. Is the software mentioned?

See the Methods section in the accompanying manuscript.

## 6a. Are alternative specifications of within-class heterogeneity considered (e.g., LGCA vs. LGMM) and clearly documented? If not, was sufficient justification provided as to eliminate certain specifications from consideration?

An LGMM which includes a varying intercept and slope by participant was used in 
this analysis and, given the nature of the within-subject variance in the data, 
is more appropriate than an LGCA.

## 6b. Are alternative specifications of the between-class differences in variance–covariance matrix structure considered and clearly documented? If not, was sufficient justification provided as to eliminate certain specifications from consideration?

The variance-covariance matrix was allowed to vary by a class-specific 
proportional parameter. The more restrictive case of restricting all classes to 
a common variance-covariance matrix structure was not considered.

## 7. Are alternative shape/functional forms of the trajectories described?

We initially modeled the trajectories using a cubic function, to allow for the 
depiction of any curvilinear decline in performance. The resulting trajectory 
in the one-class cubic model (which had the lowest BIC) was effectively flat, 
and hence we moved to linear modelling for the ease of interpretation of the 
model coefficients.

## 8. If covariates have been used, can analyses still be replicated?

Not applicable: age was the only predictor in the models.

## 9. Is information reported about the number of random start values and final iterations included?

The `gridsearch()` function of the `lcmm` package was used to run 500 departures
from random initial values for each multi-class model, using the one-class model
to generate those starting values from. The maximum number of iterations within 
the `hlme()` function was set at 1000.

## 10. Are the model comparison (and selection) tools described from a statistical perspective?

We selected the two-class linear model on the basis of it converging and having 
the lowest BIC value.

## 11. Are the total number of fitted models reported, including a one-class solution?

For both linear and cubic trajectories, models were fitted with one through five
classes. See the model comparison tables in Tables 
\@ref(tab:tab-A1-linear-model-table) and \@ref(tab:tab-A2-cubic-model-table).

(ref:tab-A1-caption) Comparison of hlme models with G = 1 to `r max_classes_linear` linear latent trajectory classes.

```{r tab-A1-linear-model-table, results='asis'}

table_a1 = linear_WTAR_model_table %>% 
  apa_table(caption = '(ref:tab-A1-caption)', 
            digits = c(0,1,0,1,2,1,1,1,1,1), align = 'r')
# only way I could find to suppress NAs (setting options(knitr.kable.NA = '') didn't work):
cat(gsub('\\bNA\\b', '  ', table_a1), sep = '\n')

```

(ref:tab-A2-caption) Comparison of hlme models with G = 1 to `r max_classes_cubic` cubic latent trajectory classes.

```{r tab-A2-cubic-model-table, results='asis'}

table_a2 = cubic_WTAR_model_table %>% 
  apa_table(caption = '(ref:tab-A2-caption)', 
            digits = c(0,1,0,1,2,1,1,1,1,1), align = 'r')
cat(gsub('\\bNA\\b', '  ', table_a2), sep = '\n')

```

## 12. Are the number of cases per class reported for each model (absolute sample size, or proportion)?

```{r include=FALSE}
# need to run this in its own chunk, with output silenced,
pp = postprob(chosen_linear_model)

```

(ref:tab-A3-caption) Number within each class in the selected model.

```{r tab-A3, results='asis'}

pp[[1]] %>% 
  apa_table(caption = '(ref:tab-A3-caption)') %>%
  gsub('.00', '', x = .) %>%
  gsub('class1', 'Typical', x = .) %>%
  gsub('class2', 'High', x = .) %>%
  cat(., sep = '\n')
```

The number of cases within each class of the selected model is shown in Table 
\@ref(tab:tab-A3). See Tables \@ref(tab:tab-A1-linear-model-table) and \@ref(tab:tab-A2-cubic-model-table) for the percentage of cases within each 
class in all candidate models.

## 13. If classification of cases in a trajectory is the goal, is entropy reported?

Classification of cases was not a goal of the analysis. However, entropy-related
measures are shown in Tables \@ref(tab:tab-A4) and \@ref(tab:tab-A5).

(ref:tab-A4-caption) Confidence of assignment to each class within the selected 
model.

```{r tab-A4, results='asis'}
pp[[2]] %>% 
  apa_table(caption = '(ref:tab-A4-caption)') %>%
  gsub('class1', 'Typical', x = .) %>%
  gsub('class2', 'High', x = .) %>%
  cat(., sep = '\n')
```

(ref:tab-A5-caption) Levels of classification confidence for each class within 
the selected model.

```{r tab-A5, results='asis'}
pp[[3]] %>% 
  apa_table(caption = '(ref:tab-A5-caption)') %>%
  gsub('class1', 'Typical', x = .) %>%
  gsub('class2', 'High', x = .) %>%
  cat(., sep = '\n')
```

## 14a. Is a plot included with the estimated mean trajectories of the final solution?

The trajectories are depicted within Figure \@ref(fig:figure-2-trajectories) of 
the accompanying manuscript, and in isolation in Figure 
\@ref(fig:fig-A3-class-trajectories).

```{r fig-A3-class-trajectories, fig.cap='The predicted trajectories of the chosen two-class linear model.', fig.height=10/2.54, fig.width=10/2.54}
# just a plot of the class lines:
figure_linear_2 = chosen_linear_predictions %>% 
  filter(ng == 2) %>% 
  ggplot(aes(x = age)) +
  geom_ribbon(aes(ymin = lower.Ypred, ymax = upper.Ypred, 
                  fill = class_for_lines), colour = NA, alpha = 0.3) +
  geom_line(aes(y = Ypred, colour = class_for_lines)) +
  geom_dl(aes(y = Ypred, label = class_for_lines),
          method = list('far.from.others.borders', cex = 0.75,
                        dl.trans(x = x + 0.30, y = y + 0.15))) +
  scale_colour_manual(values = blue_orange_sat) +
  scale_fill_manual(values = blue_orange_sat) +
  scale_y_continuous(breaks = c(80, 100, 120), limits = c(80, 130)) +
  scale_x_continuous(minor_breaks = NULL,
                     labels = scales::label_number(suffix = ' yrs')) +
  labs(title = 'A. Predicted WTAR scores by class',
       x = NULL, y = NULL) +
  theme_bw() + 
  theme(aspect.ratio = 1,
        plot.title = element_text(face = 'bold', size = 10),
        axis.ticks = element_line(colour = '#E8E8E8'),
        axis.text = element_text(size = 8),
        axis.title = element_text(size = 9),
        panel.border = element_rect(colour = '#E8E8E8', size = 1.0),
        strip.background = element_rect(fill = '#E8E8E8',
                                        colour = '#E8E8E8'),
        panel.spacing.x = unit(10, 'mm'),
        legend.position = 'none')

figure_linear_2
```

## 14b. Are plots included with the estimated mean trajectories for each model?

Figure \@ref(fig:fig-A4-all-class-trajectories) shows the predicted trajectories
of all 5 candidate linear and cubic models, as tabulated in Tables \@ref(tab:tab-A1-linear-model-table) and \@ref(tab:tab-A2-cubic-model-table).

(ref:fig-A4-all-class-trajectories) The predicted trajectories of all candidate 
latent trajectory models (the selected one was the two-class linear trajectory 
model, as shown in Figure \@ref(fig:fig-A3-class-trajectories)). The percentage 
of participants assigned to each class in each model can be found in Tables \@ref(tab:tab-A1-linear-model-table) and \@ref(tab:tab-A2-cubic-model-table).

```{r fig-A4-all-class-trajectories, fig.cap='(ref:fig-A4-all-class-trajectories)', fig.height=20/2.54, fig.width=17.5/2.54}
# just a plot of the class lines:
figure_linear_all = linear_predictions %>% 
  ggplot(aes(x = age)) +
  facet_grid(ng ~ .) +
  geom_ribbon(aes(ymin = lower.Ypred, ymax = upper.Ypred, 
                  fill = class), colour = NA, alpha = 0.3) +
  geom_line(aes(y = Ypred, colour = class)) +
  geom_dl(aes(y = Ypred, label = class), method = list('last.bumpup', cex = 0.65)) +
  scale_y_continuous(limits = c(80, 130), expand = expansion(add = 0), breaks = c(80, 100, 120)) +
  scale_x_continuous(limits = c(40, 90), breaks = c(50, 60, 70, 80, 90), 
                     expand = expansion(add = c(0, 7)), minor_breaks = NULL) +
  scale_colour_manual(values = c('#D55E00', '#0072B2', '#AAAB2A', 'brown', 'purple'), name = 'Class', guide = 'none') + 
  scale_fill_manual(values = c('#D55E00', '#0072B2', '#AAAB2A', 'brown', 'purple'), name = 'Class', guide = 'none') + 
  labs(title = 'Class trajectories for each candidate \nlinear trajectory model',
       subtitle = 'Ranging from 1 to 5 classes',
       x = 'Age', y = 'WTAR score') +
  theme_bw() + 
  theme(aspect.ratio = 0.5,
        plot.title = element_text(face = 'bold', size = 10),
        plot.subtitle = element_text(size = 9),
        axis.ticks = element_line(colour = '#E8E8E8', size = 0.25),
        axis.text = element_text(size = 8),
        axis.title = element_text(size = 9),
        panel.border = element_rect(colour = '#E8E8E8', size = 0.25),
        panel.grid.major.x = element_blank(),
        panel.grid.major.y = element_line(size = 0.25),
        strip.background = element_rect(fill = '#E8E8E8',
                                        colour = '#E8E8E8'),
        panel.spacing.x = unit(10, 'mm'),
        legend.position = 'none')

figure_cubic_all = cubic_predictions %>% 
  ggplot(aes(x = age)) +
  facet_grid(ng ~ .) +
  geom_ribbon(aes(ymin = lower.Ypred, ymax = upper.Ypred, 
                  fill = class), colour = NA, alpha = 0.3) +
  geom_line(aes(y = Ypred, colour = class)) +
  geom_dl(aes(y = Ypred, label = class), method = list('last.bumpup', cex = 0.65)) +
  scale_y_continuous(limits = c(80, 130), expand = expansion(add = 0), breaks = c(80, 100, 120)) +
  scale_x_continuous(limits = c(40, 90), breaks = c(50, 60, 70, 80, 90), 
                     expand = expansion(add = c(0, 7)), minor_breaks = NULL) +
  scale_colour_manual(values = c('#D55E00', '#0072B2', '#AAAB2A', 'brown', 'purple'), name = 'Class', guide = 'none') + 
  scale_fill_manual(values = c('#D55E00', '#0072B2', '#AAAB2A', 'brown', 'purple'), name = 'Class', guide = 'none') + 
  labs(title = 'Class trajectories for each candidate \ncubic trajectory model',
       subtitle = 'Ranging from 1 to 5 classes',
       x = 'Age', y = 'WTAR score') +
  theme_bw() + 
  theme(aspect.ratio = 0.5,
        plot.title = element_text(face = 'bold', size = 10),
        plot.subtitle = element_text(size = 9),
        axis.ticks = element_line(colour = '#E8E8E8', size = 0.25),
        axis.text = element_text(size = 8),
        axis.title = element_text(size = 9),
        panel.border = element_rect(colour = '#E8E8E8', size = 0.25),
        panel.grid.major.x = element_blank(),
        panel.grid.major.y = element_line(size = 0.25),
        strip.background = element_rect(fill = '#E8E8E8',
                                        colour = '#E8E8E8'),
        panel.spacing.x = unit(10, 'mm'),
        legend.position = 'none')

figure_A4 = cowplot::plot_grid(figure_linear_all, figure_cubic_all, ncol = 2,
                              align = 'h')
figure_A4
```

## 14c. Is a plot included of the combination of estimated means of the final model and the observed individual trajectories split out for each latent class?

This is shown in Figure \@ref(fig:figure-2-trajectories) of the accompanying 
manuscript.

## 15.Are characteristics of the final class solution numerically described (i.e., means, SD/SE, n, CI, etc.)?

Yes, see the Results section of the accompanying manuscript.

## 16. Are the syntax files available (either in the appendix, supplementary materials, or from the authors)?

Yes, they are publicly available at [github.com/nzbri/wtar-trajectory](https://github.com/nzbri/wtar-trajectory).



